---
title: "CartoDB: Using CartoDB maping tools with R"
author: "Virgilio Gómez-Rubio"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

[CartoDB](https://cartodb.com/)  is an open-source platform for cloud mapping.
In this vignette we introduce the `CartoDB` package for the R software and how
to combine both for mapping and spatial analysis.  `CartoDB` was orginially
wirtten by Andrew Hill and here I am describing a [forked version of the 
package](https://github.com/becarioprecario/cartodb-r)
that includes some other functionalities, such data, data importing, handling
of named maps, etc.

In this vignette you will learn:

- Basics about CartoDB API
- How to upload data to CartoDB from R
- ...

## Introduction

CartoDB provides limited free services through their web interface. These
includes an easy-to-use interface to upload data and a simple map inerface to
produce maps. All registered users can also access to CartoDB API to remotely
connect to CartoDB map server and perform several operations.

CartoDB provides an API Key that is required for many of the operations that we
are going to discuss in this vignette. This API Key must be kept secreat as it
would provide access to our maps and datasets to othe users.


I have included my CartoDB username (`becarioprecario`) and API Key in
variables `CBDID` and `APIKey`, respectively. They will not be defined
anywhere in this vignette (you must do so on your own) but they will
be used extensively in this vignette.

```{r, echo = FALSE}
  load("CartoDB_cred.RData")
```

Our first action is to set our credentials using the following code:

```{r, results = 'hide'}
library(CartoDB)
cartodb(CBDID, APIKey)
```

```{r, echo = FALSE}
#Delete nc_sids table
cartodb.df("DROP TABLE nc_sids")
```

Access to CartoDB API is through cURL and, for that, commands must be
sent by accessing to specific URLs that need to be built according to our
needs. Some of the functions in `CartoDB` will help us to create the
queries and construct the right URLs.

## Uploading data into CartoDB

Our first step now will be to upload a dataset into our CartoDB account.
For this, we will be using the North Carolina SIDS dataset that is
provide as a shapefile in package `spdep`. In order to load this
shapefile into R we will load some required packages first, and then
use function `readShapePoly()` to actually load the data:

```{r, results = 'hide'}
library(spdep)
library(maptools)
nc.sids <- readShapePoly(system.file("etc/shapes/sids.shp", package="spdep")[1],
       ID="FIPSNO", proj4string=CRS("+proj=longlat +ellps=clrk66"))
```

This will create a `SpatialPolygonsDataFrame` with the boundaries of the 100
counties in North Carolina and their associated data. The manual page of
`nc.sids` is where the previous code has been taken from and it also provides
more information about this dataset.

Function `r2cartodb()` (originally written by [Kyle Walker](https://rpubs.com/walkerke/r2cartodb)) will take a `Spatial*` object, export it to a shapefile
and then upload it into our CartoDB account. This function will use
the username and API Key set in call to `cartodb()`. Now, we will
upload the data in `nc.sids` into a new database called "nc.sids":

```{r, eval = TRUE}
library(rgdal)
r2cartodb(nc.sids, "nc.sids")
```

Here, the first argument is the name of the variable in R and the second
the name given to the data in our CartoDB account. Note that CartoDB
will rename the dataset to "nc_sids"

## Querying data from CartoDB

CartoDB is built on PostgreSQL and its developers praise to provide a
flexible and fast way of querying data. We will see now how to get data
from our newly updated data. In particular, we will use Transylvania county
(yes, there is a Transylvania in North Carolina but not the one you are
thinking of at the moment!!). We will build a simple query to select all
the columns in our data for Transylvania county:


```{r}
trans.data <- cartodb.df("SELECT * FROM nc_sids WHERE name = 'Transylvania'")
trans.data
```

CartoDB will return all the columns ir our original dataset plus some other
that have been added to define the geometry of the spatial data.
This will become clear once we have inspected the names:

```{r}
names(trans.data)
```

More complex queries can be build and send to the CartoDB server through its 
API.

## Spatial analysis of the data

The North Carolina SIDS data has already been analysed in multitude of research
papers and has appeared in numerous worked examples in many books.  Shortly,
this dataset includes the cases of Sudden Infant Syndrome in children in the
100 counties in North Carolina in two time periods: 1974-1978 and 1979-1984.
In addition to the counts of cases, the number of births per county (i.e.,
population at risk) and the number of non-white births (a possible confounder
or risk factor) are also available.

As a first step in the analysis, we may consider how many cases per birth we
have in North Carolina in each time period. This can be computed by dividing
the total number of cases by the total number of births, for a given period.
We will focus on the first time period. Using CartoDB API this can
be computed as:

```{r}
sql.qr <- "SELECT sum(bir74) as x from nc_sids"
tot.births74 <- cartodb.df(sql.qr)
tot.births74$x
```

Similarly, the total number of cases can be computed:

```{r}
sql.qr <- "SELECT sum(sid74) as x from nc_sids"
tot.sids74 <- cartodb.df(sql.qr)
tot.sids74$x
```

Finally, the global incidence rate for the first period can be computed as

```{r}
rate74 <- tot.sids74$x / tot.births74$x
rate74
```

Note that this requires sending queries and data back and forth between
R and CartoDB. A more ellegant solution is to compute the rate directly
on the server using a SQL query:

```{r}
sql.qr <- "SELECT sum(sid74)/sum(bir74) as x from nc_sids"
rate74.sql <- cartodb.df(sql.qr)
rate74.sql$x
```

In this case it really does not matter how we compute this rate. For large
datasets, however, it may be a good idea to make computations on the server.

The rate is better understood when it is multiplied by, for example, 1000,
because it shows the expected number of SIDS cases pero 1000 births:

```{r}
1000 * rate74.sql$x
```

This global rate will give as an indicator on the prevalence of the disease.
This can be used as a threshold to compare the county rates to. If a county
rate is higher than the global rate, then incidence in this county is higher
than average. Computing this county rates is also something that
can be done with a SQl query to CartoDB:

```{r}
sql.qr <- "SELECT sid74/bir74 from nc_sids"
county.rates <- cartodb.df(sql.qr)
summary(unlist(county.rates))
```

Another baseline indicator that can be computed for each county is the
'expected number of cases', which measures how many cases would appear in a
county if the incidence is exactly the same as in the global rate.  This can be
obtained by multipliying the global rate to the county number of births.
The advantage of this expected counts over the previous rates is that this
can be directly compared to the pbserved number of cases to assess
the prevalence of the disease.

Furthermore, the Standardised Mortality Ratio (SMR) is defined as the observed
counts divided by the expected counts. A SMR value of 1 indicates that a
county has the same observed and expected counts, i.e., the prevalence
of the disease is the same as in the whole study area. A SMR higher than one
indicates an increased prevalen, whilst a SMR lower than 1 occurs when
the prevalence is low (as compared to the hole study region).

In CartoDB, this expected counts can be computed using a SQl query as
follows:

```{r}
sql.qr <- "SELECT nc_sids.bir74 * rate.rate74 FROM nc_sids, (SELECT sum(sid74)/sum(bir74) as rate74 FROM nc_sids) rate"
exp.cases74 <- cartodb.df(sql.qr)
summary(unlist(exp.cases74))
```
Given that our aim is to compute the SMR for each county, the expected
counts can be computed and added as a new column to the table instead
of sending them back to R.

In order to make the new query we need to:

- Compute the expected number of cases
- Keep an id for each county (for example, using its name)
- Insert the new values into the table, in a new column, so that new values are matched by their ids

We will be using column `cartodb_id` as the main id for each county. The
following query will simply return the expected counts and the area name:


```{r}
sql.qr <- "SELECT nc_sids.cartodb_id, nc_sids.bir74 * rate.rate74 as exp74 FROM nc_sids, (SELECT sum(sid74)/sum(bir74) as rate74 FROM nc_sids) rate"
data.sql <- data.frame(cartodb.df(sql.qr))
data.sql[1:5, ]
```

The first step here is to add a new column to `nc_sids`:

```{r, eval = TRUE}
sql.qr <- "ALTER TABLE nc_sids ADD COLUMN exp74 numeric"
res <- cartodb.df(sql.qr)
```

The values of this new column are updated with the computed SMR's:

```{r}
sql.qr <- "UPDATE nc_sids SET exp74 = tab.exp74 FROM (SELECT nc_sids.cartodb_id, nc_sids.bir74 * rate.rate74 as exp74 FROM nc_sids, (SELECT sum(sid74)/sum(bir74) as rate74 FROM nc_sids) rate) tab WHERE nc_sids.cartodb_id = tab.cartodb_id"

res <- cartodb.df(sql.qr)
```
Now that the we have the observed and expected counts in the table, the
SMR's can be computed. As before, we will be using SQL queries to keep
all computations in the CartoDB server:

```{r}
cartodb.df("ALTER TABLE nc_sids ADD COLUMN smr74 numeric")

sql.qr <- "UPDATE nc_sids SET smr74 = tab.smr74 FROM (SELECT cartodb_id, bir74/exp74 as smr74 FROM nc_sids) tab WHERE nc_sids.cartodb_id = tab.cartodb_id"
cartodb.df(sql.qr)
```



